Explaining Black-Box Models through Counterfactuals

### 2023-11-17 examples/CounterfactualExplanations.jl/paper.qmd
this seems to work

Numerous extensions to this simple approach have been developed since CE were first proposed in 2017 (see @verma2020counterfactual and [@karimi2020survey] for surveys). The various approaches largely differ in that they use different flavours of search objective defined in @eq-solution. Different penalties are often used to address many of the desirable properties of effective CE that have been set out. These desiderata include: **proximity** --- the distance between factual and counterfactual features should be small [@wachter2017counterfactual]; **actionability** --- the proposed recourse should be actionable [@ustun2019actionable, @poyiadzi2020face]; **plausibility** --- the counterfactual explanation should be plausible to a human [@joshi2019realistic, @schut2021generating]; **sparsity** --- the counterfactual explanation should involve as few individual feature changes as possible [@schut2021generating]; **robustness** --- the counterfactual explanation should be robust to domain and model shifts [@upadhyay2021robust]; **diversity** --- ideally multiple diverse counterfactuals should be provided [@mothilal2020explaining]; and **causality** --- counterfactuals should respect the structural causal model underlying the data generating process [@karimi2020algorithmic,@karimi2021algorithmic].

@@

A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (@tbl-gen). At a high level, we distinguish generators in terms of their compatible model types, their default search space, and their composability. All "gradient-based" generators are compatible with differentiable models, e.g. `Flux` and `torch`, while "tree-based" generators are only applicable to models that involve decision trees. Concerning the search space, it is possible to search counterfactuals in a lower-dimensional latent embedding of the feature space that implicitly encodes the data-generating process (DGP). To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge [@joshi2019realistic, @karimi2021algorithmic]. While this notion is compatible with all of our gradient-based generators, only some generators search a latent space by default. Finally, composability implies that the given generator can be blended with any other composable generator, which we discuss in @sec-gen-comp.

@@

\onecolumn

@@

\twocolumn

>>>>

Numerous extensions to this simple approach have been developed since CE were first proposed in 2017 (see @verma2020counterfactual and @karimi2020survey for surveys). The various approaches largely differ in that they use different flavours of search objective defined in @eq-solution. Different penalties are often used to address many of the desirable properties of effective CE that have been set out. These desiderata include: **proximity** --- the distance between factual and counterfactual features should be small [@wachter2017counterfactual]; **actionability** --- the proposed recourse should be actionable [@ustun2019actionable; @poyiadzi2020face]; **plausibility** --- the counterfactual explanation should be plausible to a human [@joshi2019realistic;@schut2021generating]; **sparsity** --- the counterfactual explanation should involve as few individual feature changes as possible [@schut2021generating]; **robustness** --- the counterfactual explanation should be robust to domain and model shifts [@upadhyay2021robust]; **diversity** --- ideally multiple diverse counterfactuals should be provided [@mothilal2020explaining]; and **causality** --- counterfactuals should respect the structural causal model underlying the data generating process [@karimi2020algorithmic;@karimi2021algorithmic].

@@

A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (@tbl-gen). At a high level, we distinguish generators in terms of their compatible model types, their default search space, and their composability. All "gradient-based" generators are compatible with differentiable models, e.g. `Flux` and `torch`, while "tree-based" generators are only applicable to models that involve decision trees. Concerning the search space, it is possible to search counterfactuals in a lower-dimensional latent embedding of the feature space that implicitly encodes the data-generating process (DGP). To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge [@joshi2019realistic;@karimi2021algorithmic]. While this notion is compatible with all of our gradient-based generators, only some generators search a latent space by default. Finally, composability implies that the given generator can be blended with any other composable generator, which we discuss in @sec-gen-comp.

@@

@@
